---
title: "4. Introduction to ShortRead"
author: "Maruf Ahmed Bhuiyan"
date: "7/27/2020"
output:
  word_document: default
  html_document: default
---

```{r}
setwd("/Users/marufahmedbhuiyan/Desktop/DataCamp/Bioconductor")
```

# Install and Load ShortRead

```{r}
# BiocManager::install("ShortRead")

require(ShortRead)

methods(class = "ShortRead")

methods(class = "ShortReadQ")
```

# Reading in files
From the video, you've learned the difference between fasta and fastq files and what information can be stored in those files.

You have also seen examples of reading fasta and fastq files with their respective functions. You have learned that readFasta() and readFastq() both need a location and a file pattern to read one or many files.

Which two arguments are particular to the function readFastq() from the package ShortRead?

```{r}
?readFastq

# qualityType and Filter
```

# Exploring a fastq file
Fastq files usually contain thousands/millions of reads and can become very large in size! For this exercise, you will use a small fastq sub sample of 500 reads which fits easily into memory and can be read entirely using the function readFastq().

The original sequence file comes from Arabidopsis thaliana, provided by the UC Davis Genome Center. The accession number is SRR1971253 and was downloaded from the Sequence Read Archive (SRA). It contains DNA from leaf tissues, pooled and sequenced on Illumina HiSeq 2000. These sequences are single-read sequences with 50 base pairs (bp) length.

fqsample is a ShortReadQ object and contains information about reads, quality scores, and ids. It's your turn to explore it!

```{r}

# Print fqsample
fqsample <- readFastq("SRR1971253.fastq")
fqsample

# Check class of fqsample
class(fqsample)

# Check class sread fqsample
class(sread(fqsample))

# Check id of fqsample
id(fqsample)

```

# Extract a sample from a fastq file
It is your turn to draw a sample piece from a sequence of many reads.

You will use the same file you've read in the previous exercise. This file has 500 reads, each of 50 bp. The file path is stored in an object called f.

Using FastqSampler(con = file_path, n = length), set.seed(), and yield() you can subset 100 reads from your sequence file.

```{r}
# Load ShortRead
library(ShortRead)

# Set a seed for sampling
set.seed(1234)

file_path <- "/Users/marufahmedbhuiyan/Desktop/DataCamp/Bioconductor/SRR1971253.fastq"

# Use FastqSampler with f and select 100 reads
fs <- FastqSampler(con = file_path, n = 100)

# Generate new sample yield
my_sample <- yield(fs)

# Print my_sample
my_sample

# FastqSampler() draws a sub-sample from a fastq file and 
# yield() is the function used to extract the sample from the FastqSampler().
```

# Intro to Sequence Quality

```{r}
# fqsample <- readFastq("SRR1971253.fastq")

# Reads
sread(fqsample)

# Quality score of reads
quality(fqsample)

# Phred Quality score
pq <- PhredQuality(quality(fqsample))
qs <- as(pq, "IntegerList")
qs

# Quality Assessment
qa_summary <- qa(fqsample,  type = "fastq", lane = 1)
qa_summary

names(qa_summary)

# Quality report summary
browseURL(report(qa_summary))

# ASCII encoding for the Quality Scores
encoding(quality(fqsample))

alphabet(sread(fqsample))

# abc <- alphabetByCycle(alphabet(sread(fqsample)))

# nuc_by_cycle <- t(abc[1:4,])

```
# Exploring sequence quality
It is your turn to perform the quality control check on the fqsample. This is an important step, done before starting any further analyses, to quickly find out whether the data is OK or if it has any problems.

To check the encoding values for each letter in quality(), use encoding():

encoding(quality(fqsample))
For a quality assessment (QA) summary use qa():

qaSummary <- qa(fqsample, type = "fastq", lane = 1)
This qaSummary has already been created for you. QA elements can be accessed with qaSummary[["nameElement"]] where nameElement is the name of the element you wish to inspect.

```{r}

# Check quality
quality(fqsample)

# Check encoding of quality
encoding(quality(fqsample))

# Check baseQuality
qa_summary[["baseQuality"]]

# Create a html report
browseURL(report(qa_summary))
```

# Try your own nucleotide frequency plot
Now it's time to take a closer look at the frequency of nucleotides per cycle. The best way to do this is by making a visualization. Usually, the first cycles are a bit random and then the frequency of nucleotides should stabilize with the coming cycles.

This exercise uses the complete fastq file SRR1971253 with some pre-processing done for you:

library(ShortRead)
fqsample <- readFastq(dirPath = "data", 
                      pattern = "SRR1971253.fastq")
# extract reads                      
abc <- alphabetByCycle(sread(fqsample))

# Transpose nucleotides A, C, G, T per column
nucByCycle <- t(abc[1:4,]) 

# Tidy dataset
nucByCycle <- nucByCycle %>% 
  as.tibble() %>% # convert to tibble
  mutate(cycle = 1:50) # add cycle numbers
Your task is to make a Nucleotide Frequency by Cycle plot using tidyverse functions!

```{r}

fqsample <- readFastq(dirPath = getwd(), 
                      pattern = "SRR1971253.fastq")
fqsample

# extract reads                      
abc <- alphabetByCycle(sread(fqsample))
abc

# Transpose nucleotides A, C, G, T per column
nucByCycle <- t(abc[1:4,]) 
nucByCycle

library(tidyverse)

# Tidy dataset
nucByCycle <- nucByCycle %>% 
  as_tibble() %>% # convert to tibble
  mutate(cycle = 1:50) # add cycle numbers

nucByCycle 

# ?gather

# Glimpse nucByCycle
glimpse(nucByCycle)

# Create a line plot of cycle vs count
nucByCycle %>% 
  # Gather the nucleotide letters in alphabet and get a new count column
  pivot_longer(-cycle, names_to = "alphabet", values_to = "count") %>%
  # gather(key = alphabet, value = count , -cycle) %>% 
  ggplot(aes(x = cycle, y =  count, color = alphabet)) +
  geom_line(size = 1 ) +
  labs(y = "Frequency") +
  theme_bw() +
  theme(panel.grid.major.x = element_blank())
  
# vignette("pivot")

```
# Math & Filter
Filtering reads on the go!
What if, from all of the reads in a file, you are only interested in some of those reads? You can use a filter!

Let's say that you are interested only in those reads that start with the pattern "ATGCA". A tiny filtering function can do the job, making use of the srFilter() function:

myStartFilter <- srFilter(function(x) substr(sread(x), 1, 5) == "ATGCA")
This function, which has been created for you, takes a ShortRead derived object as an input and outputs the reads starting with the pattern "ATGCA". Let's put this function to use!

```{r}
# Check class of fqsample
class(fqsample)

myStartFilter <- srFilter(function(x) 
                substr(sread(x), 1, 5) == "ATGCA")

# Filter reads into selectedReads using myStartFilter
selectedReads <- fqsample[myStartFilter(fqsample)]

# Check class of selectedReads
class(selectedReads)

# Check detail of selectedReads
detail(selectedReads)
```

# Removing duplicates
It is always a good practice to check that your sequence reads don't contain too many duplicates.

# Sample with duplicates of class: ShortReadQ
dfqsample

# Get the reads from dfqsample
mydReads <- sread(dfqsample)

# Counting duplicates
table(srduplicated(mydReads))
How would you go about filtering duplicated reads in a file? Pay attention to what the condition should be in this filter.

```{r}
# mydReads[srduplicated(mydReads) == FALSE]

# To keep those reads that are not duplicated 
# mydReads[srduplicated(mydReads) == FALSE]
```

# More filtering!
Awesome! Now that you've had some practice with filtering reads, let's use the function polynFilter(). This function selects reads that contain less than a given number of duplicate nucleotides. For example, polynFilter(threshold = 20, nuc = c("A")) will select all reads that contain less than 20 A's. The parameter nuc is a character vector containing IUPAC symbols for nucleotides or the value "other" for all non-nucleotide symbols.

The fqsample object is available in your workspace.

```{r}
# Check reads of fqsample
sread(fqsample)

# Create myFil using polynFilter
myFil  <- polynFilter(threshold = 3, nuc = c("A"))

# Check myFil
myFil

# Check reads of fqsample
sread(fqsample)

# Create myFil using polynFilter
myFil <- polynFilter(threshold = 3, nuc = c("A"))
myFil

# Apply your filter to fqsample
filterCondition <- myFil(fqsample)
filterCondition

# Use myFil with fqsample
filteredSequences <- fqsample[filterCondition]
filteredSequences

# Check reads of filteredSequences
sread(filteredSequences)

# Compare before(fqsample) & after(filteredSeqeuences)
sread(fqsample)

# 717 reads is what we have left after applying the 
# polynFilter() and filtering using the filterCondition
# on fqsample. Once you are comfortable using conditions
# directly as subsets, you can create the object filteredSequences
# in one line of code, like this: 
# objectToFilter[customFilter(objectToFilter)]

oneline_filter <- fqsample[myFil(fqsample)]
sread(oneline_filter)

```

# Rqc

```{r}
# BiocManager::install("Rqc")

require(Rqc)
```

# Plotting cycle average quality
Here you are! To challenge yourself, are you ready to test a bit of all of what you have learned by making three quick plots? You will see that now it all pays off!

It is your turn to use the new package Rqc and check multiple files at once.

```{r}
# a <- rqcCycleAverageQualityPlot(qa)

# Average per cycle quality plot with white background
# a + theme_minimal()

# Read quality plot with white background
# a + theme_minimal()

```

















